---
title: "hvCpG_BayesAlgo_exploration"
author: "Alice Balard"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(data.table)
library(matrixStats)
library(ggplot2)
library(reshape2)
library(ggrepel)
```

# Data prep 

```{r}
##########################################################################################
## LSHTM: Maria’s data that she used for the hvCpG paper can be accessed from ing-p5 here:
# /mnt/old_user_accounts/p3/maria/PhD/Data/datasets/
#   
# Her hvCpG scripts are here:
# /mnt/old_user_accounts/p3/maria/PhD/Projects/hvCpGs/Scripts/

source("/home/alice/2024_hvCpG/03_prepDatasetsMaria/dataprep_MariaArrays.R")
my_list_mat <- Maria_filtered_list_mat; rm(Maria_filtered_list_mat)
cpgnames <- unique(unlist(sapply(my_list_mat, row.names)))
cpgnames <- cpgnames[order(cpgnames)]
```

# Reproduce Maria’s results
Maria’s paper: We defined an hvCpG in the following way:

1. in 65% of datasets in which the CpG is covered (following quality control), it has methylation variance in the top 5% of all (non-removed) CpGs.
2. is covered in at least 15 of the 30 datasets.

```{r Maria}
## Within each dataset, calculate the CpGs variance, and keep the top 5%
top5pcvar <- lapply(my_list_mat, function(mat) {
  # Step 1: Calculate row variances
  row_variances = rowVars(mat, na.rm = T)
  df = data.frame(var=row_variances, cpg=names(row_variances))
  # Step 2
  top = top_n(df, as.integer(0.05*nrow(df)), var) ## slightly different than quantile cause keep ties
  return(top)
})

## CpGs in the top 5% variance:
cpg_counts_top <- table(unlist(lapply(top5pcvar, rownames))) %>%
  data.frame() %>% dplyr::rename("cpgs"="Var1", "all_cpgs_top5pc"="Freq")

## all covered CpGs:
cpg_counts <- table(unlist(lapply(my_list_mat, rownames))) %>%
  data.frame() %>% dplyr::rename("cpgs"="Var1", "all_cpgs"="Freq")

## Both:
cpg_counts_full <- merge(cpg_counts, cpg_counts_top, all.x = T)
rm(cpg_counts, cpg_counts_top)

## in 65% of datasets in which the CpG is covered (following quality control), it has methylation variance in the top 5% of all (non-removed) CpGs:
hvCpGs_maria <- na.omit(cpg_counts_full[cpg_counts_full$all_cpgs_top5pc/cpg_counts_full$all_cpgs >= 0.65,])

## rounding, to mimick Maria's approach
hvCpGs_maria <- na.omit(cpg_counts_full[round(cpg_counts_full$all_cpgs_top5pc/cpg_counts_full$all_cpgs, 2) >= 0.65,])
```

Maria detected `r length(MariasCpGs$CpG)` hvCpGs. I detected `r length(hvCpGs_maria$cpgs)` hvCpGs.
We both have `r length(intersect(hvCpGs_maria$cpgs, MariasCpGs$CpG))` hvCpGs in common.

# Identify hvCpGs based on a sd multiplicative factor lambda

The proportion of CpGs than Maria found hvCpGs is p(hvCpG) = `r round(4143/406306, 4)*100`%.

## Calculate lambda 

Within each dataset k, calculate the median sd of all CpG j

```{r calcLambdas}
all_sd_jk <- sapply(my_list_mat, function(k){
  
  ## Scale
  k = log2(k/(1-k))
  
  get_sd_k <- function(k){
    ## Calculate a vector of the row (=per CpG j) sd
    return(rowSds(k, na.rm = T))
  }
  ## Return a vector of sds, in a list for each dataset 
  return(get_sd_k(k))
})

## Plot:
df_long <- reshape2::melt(all_sd_jk, variable.name = "Vector", value.name = "SDs")

## Plot distributions of all vectors on the same graph
ggplot(df_long, aes(x = SDs, color = L1)) +
  geom_density() +
  labs(title = "Distribution of SDs accross datasets",
       x = "SDs",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_sqrt()

## Calculate lambda per dataset
lambdas = sapply(all_sd_jk, function(x){quantile(x, 0.95, na.rm=T)/median(x, na.rm=T)})
names(lambdas) <- gsub(".95%", "", names(lambdas))
```

```{r}
# Histogram with kernel density
ggplot(data.frame(lambda=lambdas, dataset=names(lambdas)),
       aes(x = lambdas)) +
  geom_histogram(aes(y = after_stat(density)),
                 colour = 1, fill = "white", binwidth = .1) +
  geom_density(lwd = 1, colour = 4,
               fill = 4, alpha = 0.25)+
  theme_minimal() +
  geom_vline(xintercept = median(lambdas), col = "red")

median(lambdas) #1.878303
```

Which datasets have a higher lambda, and why? 

```{r}
df = data.frame(nind=sapply(my_list_mat, ncol),
                dataset=names(my_list_mat),
                lambda=lambdas, 
                tissue = sapply(strsplit(names(my_list_mat), "_"), `[`, 1),
                ethnicity=sapply(strsplit(names(my_list_mat), "_"), `[`, 2)) %>%
  mutate(dataset=ifelse(dataset %in% c("Blood_Cauc", "Blood_Hisp"), "Blood_Cauc_Hisp", dataset)) %>% 
  mutate(dataset=ifelse(dataset %in% c("Blood_Mexican", "Blood_PuertoRican "), "Blood_Mex_PuertoRican ", dataset)) %>% 
  mutate(dataset=ifelse(dataset %in% c("CD4+_Estonian", "CD8+_Estonian"), "CD4+_CD8+_Estonian", dataset)) %>% 
  mutate(dataset=ifelse(dataset %in% c("Saliva_Hisp", "Saliva_Cauc"), "Saliva_Hisp_Cauc", dataset))

ggplot(data = df,
       aes(x=lambda, y=nind))+
  geom_smooth(method = "lm")+
  geom_point()+
  geom_label_repel(aes(label = dataset, fill=dataset), size= 2, alpha=.8, max.overlaps = 25)+
  theme_bw()+theme(legend.position = "none")+
  scale_x_log10()
```

```{r}
## Emphasize the outliers
df_long$col = "grey"
df_long$col[df_long$L1 %in% "BulkFrontalCortex"] <- "red"

ggplot(df_long, aes(x = SDs, group = L1, fill = col)) +
  geom_density(data = df_long[!df_long$L1 %in% "BulkFrontalCortex",], alpha = .5) +
  geom_density(data = df_long[df_long$L1 %in% "BulkFrontalCortex",], alpha = .6) +
  scale_fill_manual(values = c("grey", "red"))+
  labs(title = "Distribution of SDs accross datasets",subtitle = "Red=BulkFrontalCortex",
       x = "SDs",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_sqrt()
```

```{r}
## Emphasize the outliers
df_long$col = "grey"
df_long$col[df_long$L1 %in% "Blood_Cauc"] <- "red"

ggplot(df_long, aes(x = SDs, group = L1, fill = col)) +
  geom_density(data = df_long[!df_long$L1 %in% "Blood_Cauc",], alpha = .5) +
  geom_density(data = df_long[df_long$L1 %in% "Blood_Cauc",], alpha = .6) +
  scale_fill_manual(values = c("grey", "red"))+
  labs(title = "Distribution of SDs accross datasets",subtitle = "Red=Blood_Cauc",
       x = "SDs",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_sqrt()
```

```{r}
## Emphasize the outlier
df_long$col = "grey"
df_long$col[df_long$L1 %in% "Blood_Hisp"] <- "red"

ggplot(df_long, aes(x = SDs, group = L1, fill = col)) +
  geom_density(data = df_long[!df_long$L1 %in% "Blood_Hisp",], alpha = .5) +
  geom_density(data = df_long[df_long$L1 %in% "Blood_Hisp",], alpha = .6) +
  scale_fill_manual(values = c("grey", "red"))+
  labs(title = "Distribution of SDs accross datasets",subtitle = "Red=Blood_Hisp",
       x = "SDs",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_sqrt()
```

# Maximum likelihood analysis

The equation is:

$$
\log\left(P(M_j)\right) = \sum_{i=1}^{n} \log\left( \sum_{Z_j=0}^{1} \left( \sum_{Z_{j,k}=0}^{1} P(M_{i,j} \mid Z_{j,k}) \times P(Z_{j,k} \mid Z_j) \right) \times P(Z_j) \right)
$$

## Examine weird values 

The transformation lead to weird values:

```{r investigate_weird_transfo}
# my_list_mat$Blood_Hisp[rownames(my_list_mat$Blood_Hisp) %in% "cg23089912",1:10]
# scaled_list_mat$Blood_Hisp[rownames(scaled_list_mat$Blood_Hisp) %in% "cg23089912",1:10]
# 
# test <- readRDS("/mnt/old_user_accounts/p3/maria/PhD/Data/datasets/GEO/Raw_cleaned_beta_matrices_GEO/Blood_Hisp")
# test[rownames(test) %in% "cg23089912",5:10]
# 
# test <- readRDS("/mnt/old_user_accounts/p3/maria/PhD/Data/datasets/GEO/BMIQ + 10 PCs + age + sex OUTLIERS REMOVED/Blood_Hisp.RDS")
# test[rownames(test) %in% "cg23089912",5:10]
# # GSM1870986
# # 0.00000000000000001124546 --> give extreme value in scaling, and p dnorm=0
```

The zero in 8h sample was weirdly transformed. The cleaned values are very different from the original. Is that expected? 

## Optimisation over multiple CpGs

Run on LSHTM server (need to be outside of RStudio): 
source("run_hvCpGdetection.R") ## for different parameters

### "Nelder-Mead" or "L-BFGS-B" optimisation method?

```{r}
load("/home/alice/2024_hvCpG/05_hvCpGalgorithm/resultsDir/results_NM_test3000_0.95.RDA")
results_NM_test3000_0.95 <- result %>% data.frame %>% tibble::rownames_to_column("CpG") %>% 
  dplyr::rename(alpha_NM_0.95=".")

load("/home/alice/2024_hvCpG/05_hvCpGalgorithm/resultsDir/results_LB_test3000_0.95.RDA")
results_LB_test3000_0.95 <- result %>% data.frame %>% tibble::rownames_to_column("CpG") %>% 
  dplyr::rename(alpha_LB_0.95="alpha")

merge(results_NM_test3000_0.95, results_LB_test3000_0.95, by = "CpG") -> results_test3000_0.95
results_test3000_0.95 %>% mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                                           "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% ggplot(aes(x=alpha_LB_0.95, y=alpha_NM_0.95, fill = Previous.results)) +
  geom_abline(slope = 1) +
  geom_point(pch=21, size = 3, alpha = .8) +
  theme_bw()
```

Some CpGs have an alpha of 0 for LBB algorithm and different alpha for NM

```{r}
results_test3000_0.95[round(results_test3000_0.95$alpha_NM_0.95, 1) != round(results_test3000_0.95$alpha_LB_0.95, 1),] %>% mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% melt() %>%
  ggplot(aes(x=CpG, y=value, fill = variable)) +
  geom_point(pch=21, size = 3, alpha = .8) +
  facet_grid(Previous.results~.) + theme_bw() + theme(axis.text.x = element_blank())
```

The probability of zero seems to be an artefact of the "L-BFGS-B" optimisation method, so we choose "Nelder-Mead".

## Test different p0

```{r}
load("/home/alice/2024_hvCpG/05_hvCpGalgorithm/resultsDir/results_NM_test3000_0.80.RDA")
results_NM_test3000_0.80 <- result %>% data.frame %>% tibble::rownames_to_column("CpG") %>% 
  dplyr::rename(alpha_NM_0.80=".")

load("/home/alice/2024_hvCpG/05_hvCpGalgorithm/resultsDir/results_NM_test3000_0.99.RDA")
results_NM_test3000_0.99 <- result %>% data.frame %>% tibble::rownames_to_column("CpG") %>% 
  dplyr::rename(alpha_NM_0.99=".")

merge(merge(results_NM_test3000_0.80, results_NM_test3000_0.99), results_NM_test3000_0.95) %>%
  mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                   "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% 
  melt() %>%
  mutate(p0 = sub(".*_", "", variable)) %>%
  ggplot(aes(x=CpG, y=value, fill = Previous.results)) +
  geom_point(pch=21, size = 2, alpha = .5) + 
  scale_fill_manual(values = c("pink","black")) +
  ylab("alpha (probability of being a hvCpG)") +
  facet_grid(.~p0) + theme_bw() + theme(axis.text.x = element_blank()) 

merge(merge(results_NM_test3000_0.80, results_NM_test3000_0.99), results_NM_test3000_0.95) %>%
  mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                   "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% 
  melt() %>%
  mutate(p0 = sub(".*_", "", variable)) %>%
  ggplot(aes(x = factor(p0), y = value, fill = Previous.results)) +
  geom_point(
    pch = 21, alpha = .2,
    position = position_jitterdodge(
      jitter.width = 0.2,  # Horizontal jitter amount
      dodge.width = 0.8     # Must match boxplot dodge width
    )
  ) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = .5) +
  scale_fill_manual(values = c("pink", "grey")) +
  xlab("p0 (probability of true negative)") +
  ylab("alpha (probability of being a hvCpG)") +
  theme_bw()
```

P0=0.95 sounds fine, not much improvement by taking 0.99

## What are the hvCpGs of Maria that I found with a very low probability of being a hvCpG with my approach?

```{r}
source("../03_prepDatasetsMaria/dataprep_MariaArrays.R")

getplotfewCpG <-function(myvec, title){
  listToTest <- lapply(Maria_filtered_list_mat, function(df) df[row.names(df) %in% myvec, , drop = FALSE])
  # Combine the list into one data frame
  mylist <- lapply(lapply(listToTest, as.data.frame), tibble::rownames_to_column, var = "CpG")
  # Combine all into one long data frame
  long_df <- bind_rows(mylist, .id = "Dataset")
  # Pivot to long format: one row per CpG/sample/tissue
  df_long <- na.omit(long_df %>%
                       tidyr::pivot_longer(
                         cols = -c(Dataset, CpG),
                         names_to = "Sample",
                         values_to = "Value"
                       ))
  
  # Plot with color by source data frame
  ggplot(df_long, aes(x = Dataset, y = Value, color = Dataset)) +
    geom_point() +
    theme_bw() + theme(legend.position = "none") + facet_grid(.~CpG) +
    ggtitle(title)+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5))  
}

myvec1 <- results_NM_test3000_0.95 %>%
  mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                   "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% 
  arrange(., alpha_NM_0.95) %>%
  filter(Previous.results %in% "1500 hvCpG in Maria's study") %>% head(n=3) %>% select(CpG) %>%
  unlist()
getplotfewCpG(myvec1, "Maria's hvCpG with the 3 lowest probabilities to be one in Alice's")

myvec2 <- results_NM_test3000_0.95 %>%
  mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                   "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% 
  arrange(., desc(alpha_NM_0.95)) %>%
  filter(Previous.results %in% "1500 hvCpG in Maria's study") %>% head(n=3) %>% select(CpG) %>%
  unlist()
getplotfewCpG(myvec2, "Maria's hvCpG with the 3 highest probabilities to be one in Alice's")

myvec3 <- results_NM_test3000_0.95 %>%
  mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                   "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% 
  arrange(., desc(alpha_NM_0.95)) %>%
  filter(!Previous.results %in% "1500 hvCpG in Maria's study") %>% head(n=3) %>% select(CpG) %>%
  unlist()
getplotfewCpG(myvec3, "CpG not detected by Maria as hvCpG, with the 3 highest probabilities to be one in Alice's")

myvec4 <- results_NM_test3000_0.95 %>%
  mutate(Previous.results = ifelse(CpG %in% MariasCpGs$CpG, 
                                   "1500 hvCpG in Maria's study", "1500 mQTL matching controls")) %>% 
  arrange(., alpha_NM_0.95) %>%
  filter(!Previous.results %in% "1500 hvCpG in Maria's study") %>% head(n=3) %>% select(CpG) %>%
  unlist()
getplotfewCpG(myvec4, "CpG not detected by Maria as hvCpG, with the 3 lowest probabilities to be one in Alice's")
```

# Notes:

lambdas are defined based on a 5% threshold, even if alpha is not. How to not hardcode them? MCMC on lambda? Link with alpha?

# The analysis is complete.
