---
title: "hvCpG_BayesAlgo_exploration"
author: "Alice Balard"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup}
library(here)

knitr::opts_chunk$set(echo = TRUE)
invisible(lapply(
  c("dplyr", "data.table", "matrixStats", "ggplot2", "reshape2",
    "ggrepel", "pROC", "stringr", "viridis", "ggridges"),
  function(pkg) suppressPackageStartupMessages(library(pkg, character.only = TRUE))
))
```

# Data prep 

```{r}
##########################################################################################
## LSHTM: Maria’s data that she used for the hvCpG paper can be accessed from ing-p5 here:
# /mnt/sod/old_user_accounts/p3/maria/PhD/Data/datasets/
#   
# Her hvCpG scripts are here:
# /mnt/sod/old_user_accounts/p3/maria/PhD/Projects/hvCpGs/Scripts/

## Load results of Maria for hvCpG
Marias_hvCpGs <- read.csv("~/2024_hvCpG/B_MultiTissues/01_dataPrep/prepDatasetsMaria_LSHTMserver/Derakhshan2022_ST5_hvCpG.txt")

## Recreate Maria's datasets
folder1 <- normalizePath("/mnt/sod/old_user_accounts/p3/maria/PhD/Data/datasets/GEO/BMIQ + 10 PCs + age + sex OUTLIERS REMOVED/")
rds_files1 <- list.files(path = folder1, pattern = "\\.RDS$", full.names = TRUE)

# Read all .rds files into a list and convert each to a matrix
rds_list_mat1 <- lapply(rds_files1, function(file) {as.matrix(readRDS(file))})
## Name the list elements by file names (without extensions)
names(rds_list_mat1) <- gsub("\\.RDS$", "", basename(rds_files1))

rds_list_mat2 <- readRDS("/mnt/sod/old_user_accounts/p3/maria/PhD/Data/datasets/TCGA/TCGA_BMIQ_age_sex_PC_adjusted_OUTLIERS_REMOVED_round2.RDS")
rds_list_mat2 <- lapply(rds_list_mat2, as.matrix)

rds_list_mat <- c(rds_list_mat1, rds_list_mat2) ; rm(rds_list_mat1, rds_list_mat2)

# Keep only the array background (covered in 15 datasets out of 30)
all_cpgs <- unlist(lapply(rds_list_mat, rownames))
cpg_counts <- table(all_cpgs)
common_cpgs <- names(cpg_counts[cpg_counts >= 15])

rds_list_mat <- lapply(rds_list_mat, function(mat) {
  mat[rownames(mat) %in% common_cpgs, ]
})

## Load mQTL-matched controls
cistrans_GoDMC_hvCpG_matched_control <- 
  read.table("~/2024_hvCpG/B_MultiTissues/01_dataPrep/prepDatasetsMaria_LSHTMserver/cistrans_GoDMC_hvCpG_matched_control.txt", header = T)

## Load cpgnames (all cpg covered after filtration)
cpg_names_all <- rhdf5::h5read("/home/alice/arraysh5/all_matrix_noscale.h5", "cpg_names")

## Check if I'm consistent
table(cpg_names_all %in% common_cpgs) ## PERFECT

sub_cistrans_GoDMC_hvCpG_matched_control <- cistrans_GoDMC_hvCpG_matched_control[
  cistrans_GoDMC_hvCpG_matched_control$hvCpG_name %in% cpg_names_all &
    cistrans_GoDMC_hvCpG_matched_control$controlCpG_name %in% cpg_names_all,]

## Histogram of samples in array data
df <- data.frame(n_samples = sapply(rds_list_mat, ncol))

# Plot with ggplot2
ggplot(df, aes(x = n_samples)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of number of samples per dataset",
    x = "Number of samples",
    y = "Count of datasets"
  )

df

rm(all_cpgs, cpg_counts,folder1, rds_files1, cistrans_GoDMC_hvCpG_matched_control)
```

## Distribution of methylation values for array after transformation

```{r}
# combine everything into one vector
# all_values <- unlist(rds_list_mat, use.names = FALSE)
# 
# # plot one histogram for all values
# hist(all_values, breaks = 100,
#      main = "Histogram of all values in rds_list_mat",
#      xlab = "Values")

# rm(all_values)
```

# Reproduce Maria’s results
Maria’s paper: We defined an hvCpG in the following way:

1. in 65% of datasets in which the CpG is covered (following quality control), it has methylation variance in the top 5% of all (non-removed) CpGs.
2. is covered in at least 15 of the 30 datasets.

```{r Maria}
detecthvCpGCutoff <- function(rds_list_mat){
  ## Within each dataset, calculate the CpGs variance, and keep the top 5%
  top5pcvar <- lapply(rds_list_mat, function(mat) {
    # Step 1: Calculate row variances
    row_variances = rowVars(mat, na.rm = T)
    df = data.frame(var=row_variances, cpg=names(row_variances))
    # Step 2
    top = top_n(df, as.integer(0.05*nrow(df)), var) ## slightly different than quantile cause keep ties
    return(top)
  })
  
  ## CpGs in the top 5% variance:
  cpg_counts_top <- table(unlist(lapply(top5pcvar, rownames))) %>%
    data.frame() %>% dplyr::rename("cpgs"="Var1", "all_cpgs_top5pc"="Freq")
  
  ## all covered CpGs:
  cpg_counts <- table(unlist(lapply(rds_list_mat, rownames))) %>%
    data.frame() %>% dplyr::rename("cpgs"="Var1", "all_cpgs"="Freq")
  
  ## Both:
  cpg_counts_full <- merge(cpg_counts, cpg_counts_top, all.x = T)
  rm(cpg_counts, cpg_counts_top)
  
  ## in 65% of datasets in which the CpG is covered (following quality control), it has methylation variance in the top 5% of all (non-removed) CpGs:
  ## rounding, to mimick Maria's approach
  hvCpGs_repro_maria <- na.omit(cpg_counts_full[round(cpg_counts_full$all_cpgs_top5pc/cpg_counts_full$all_cpgs, 2) >= 0.65,])
  
  return(hvCpGs_repro_maria)
}

hvCpGs_repro_maria <- detecthvCpGCutoff(rds_list_mat = rds_list_mat)
```

Maria detected `r length(Marias_hvCpGs$CpG)` hvCpGs. I detected `r length(hvCpGs_repro_maria$cpgs)` hvCpGs.
We both have `r length(intersect(hvCpGs_repro_maria$cpgs, Marias_hvCpGs$CpG))` hvCpGs in common.

## What if we had only 3 people per dataset?

```{r}
set.seed(123)
rds_list_mat_3ind <- lapply(rds_list_mat, function(df) df[, sample(ncol(df), 3), drop = FALSE])
rds_list_mat_2ind <- lapply(rds_list_mat, function(df) df[, sample(ncol(df), 2), drop = FALSE])

hvCpGs_repro_maria_3ind <- detecthvCpGCutoff(rds_list_mat = rds_list_mat_3ind)
hvCpGs_repro_maria_2ind <- detecthvCpGCutoff(rds_list_mat = rds_list_mat_2ind)

x <- list(
  "Full array"     = hvCpGs_repro_maria$cpgs,
  "Array 2 ind/ds" = hvCpGs_repro_maria_2ind$cpg,
  "Array 3 ind/ds" = hvCpGs_repro_maria_3ind$cpg
)

saveRDS(x, here("B_MultiTissues/dataOut/arrayCutoffLowPower2or3ind.RDS"))
## Plotted in S02
```


# Identify hvCpGs based on a sd multiplicative factor lambda

The proportion of CpGs than Maria found hvCpGs is p(hvCpG) = `r round(4143/406306, 4)*100`%.

## Calculate lambda 

Within each dataset k, calculate the median sd of all CpG j

```{r calcLambdas}
all_sd_jk <- sapply(rds_list_mat, function(k){
  
  get_sd_k <- function(k){
    ## Calculate a vector of the row (=per CpG j) sd
    return(rowSds(k, na.rm = T))
  }
  ## Return a vector of sds, in a list for each dataset 
  return(get_sd_k(k))
})

## Plot:
df_long <- reshape2::melt(all_sd_jk, variable.name = "Vector", value.name = "SDs")

## Plot distributions of all vectors on the same graph
ggplot(df_long, aes(x = SDs, color = L1)) +
  geom_density() +
  labs(title = "Distribution of SDs accross datasets",
       x = "SDs",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_sqrt()

## Calculate lambda per dataset
lambdas = sapply(all_sd_jk, function(x){quantile(x, 0.95, na.rm=T)/median(x, na.rm=T)})
names(lambdas) <- gsub(".95%", "", names(lambdas))
```

```{r}
# Histogram with kernel density
ggplot(data.frame(lambda=lambdas, dataset=names(lambdas)),
       aes(x = lambdas)) +
  geom_histogram(aes(y = after_stat(density)),
                 colour = 1, fill = "white", binwidth = .1) +
  geom_density(lwd = 1, colour = 4,
               fill = 4, alpha = 0.25)+
  theme_minimal() +
  geom_vline(xintercept = median(lambdas), col = "red")

median(lambdas)
```

Which datasets have a higher lambda, and why? 

```{r}
df = data.frame(nind=sapply(rds_list_mat, ncol),
                dataset=names(rds_list_mat),
                lambda=lambdas, 
                tissue = sapply(strsplit(names(rds_list_mat), "_"), `[`, 1),
                ethnicity=sapply(strsplit(names(rds_list_mat), "_"), `[`, 2)) %>%
  mutate(dataset=ifelse(dataset %in% c("Blood_Cauc", "Blood_Hisp"), "Blood_Cauc_Hisp", dataset)) %>% 
  mutate(dataset=ifelse(dataset %in% c("Blood_Mexican", "Blood_PuertoRican "), "Blood_Mex_PuertoRican ", dataset)) %>% 
  mutate(dataset=ifelse(dataset %in% c("CD4+_Estonian", "CD8+_Estonian"), "CD4+_CD8+_Estonian", dataset)) %>% 
  mutate(dataset=ifelse(dataset %in% c("Saliva_Hisp", "Saliva_Cauc"), "Saliva_Hisp_Cauc", dataset))

ggplot(data = df,
       aes(x=lambda, y=nind))+
  geom_smooth(method = "lm")+
  geom_point()+
  geom_label_repel(aes(label = dataset, fill=dataset), size= 2, alpha=.8, max.overlaps = 25)+
  theme_bw()+theme(legend.position = "none")+
  scale_x_log10()
```

```{r}
## Emphasize the outliers
df_long$col = "grey"
df_long$col[df_long$L1 %in% "uterus"] <- "red"

ggplot(df_long, aes(x = SDs, group = L1, fill = col)) +
  geom_density(data = df_long[!df_long$L1 %in% "uterus",], alpha = .5) +
  geom_density(data = df_long[df_long$L1 %in% "uterus",], alpha = .6) +
  scale_fill_manual(values = c("grey", "red"))+
  labs(title = "Distribution of SDs accross datasets",subtitle = "Red=uterus",
       x = "SDs",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_sqrt()
```

# Maximum likelihood analysis

The equation is:

$$
\log\left(P(M_j)\right) = \sum_{i=1}^{n} \log\left( \sum_{Z_j=0}^{1} \left( \sum_{Z_{j,k}=0}^{1} P(M_{i,j} \mid Z_{j,k}) \times P(Z_{j,k} \mid Z_j) \right) \times P(Z_j) \right)
$$

## Examine weird values 

The transformation lead to weird values:

```{r investigate_weird_transfo}
test <- readRDS("/mnt/sod/old_user_accounts/p3/maria/PhD/Data/datasets/GEO/Raw_cleaned_beta_matrices_GEO/Blood_Hisp")
test[rownames(test) %in% "cg23089912",5:10]

test <- readRDS("/mnt/sod/old_user_accounts/p3/maria/PhD/Data/datasets/GEO/BMIQ + 10 PCs + age + sex OUTLIERS REMOVED/Blood_Hisp.RDS")
test[rownames(test) %in% "cg23089912",5:10]
# GSM1870986
# 0.00000000000000001124546 --> give extreme value in scaling, and p dnorm=0
```

The zero was weirdly transformed. Is that expected? 

# The analysis is complete.
